---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gestage

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/ki-tools/gestage/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ki-tools/gestage/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `gestage` R package implements methods described in [this paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8438948/) to estimate gestational age.

## Installation

You can install the development version of `gestage` from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("ki-tools/gestage")
```

## Example

Currently only one of the models in the paper has been implemented, "Model C", which takes as input gestational age at birth based on last menstrual period (LMP), `gagelmp`, and birth weight, `birthwt`.

```{r example}
library(gestage)

# predict GA for one set of predictors:
method_c_predict(gagelmp = 250, birthwt = 2800)

# predict GA for different values of gagelmp
method_c_predict(gagelmp = 201:225, birthwt = 2800)

# predict GA for different values of birthwt
method_c_predict(gagelmp = 250, birthwt = seq(3005, 3100, by = 5))
```
